### 2020-10-12 manuscript/manuscript.rmd
addressed feedback from Nicolas

Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Antonis Rokas^3^ * and Casey W. Dunn^1^*

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals, was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1D). Since then, many more relevant phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. This is because methodological variance is often confounded, for example when a particular model is only available in one software package.

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister. Moreover, we found that the branch lengths of Ctenophora taxa are generally longer than other animals with both site-homogeneous and site-heterogeneous models in many data matrices (*e.g.* Extended Data Figs. 3, 4).

@@

There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that site can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many categories. That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 5A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 5A-B). This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.

@@

Taken together, although further analysis are warranted, this suggests that unconstrained CAT analyses might be overparameterized. While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are problems with unconstrained CAT analyses that are prone to create a large number of categories for a small fraction of sites. In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of categories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.

@@

Note that this data matrix has been extensively reanalyzed in many studies although there is very little ctenophore sampling and Myxozoans contain extremely long-branch in the tree.

>>>>

Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Casey W. Dunn^1^* and Antonis Rokas^3^*

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals, was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1D). Since then, many more relevant phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. This is because methodological variance is often confounded, for example when a particular model is only available in one software package. There are a variety of things that make this a particularly challenging problem. For one, the relevant nodes are the deepest in the animal tree of life. Other factors that have been invoked include concern that Ctenophora-sister could be an artefact of long branch attraction to outgroups [@pisani2015genomic]. While some sponges have a longer root to tip length, indicating an accelarated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs. 3, 4), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

@@

There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that category can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many categories. That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 5A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 5A-B). This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.

@@

@@
